Aquifer depletion exacerbates agricultural drought losses in the US High Plains

### 2023-11-04 Writing/nw_main.Rnw
a code to split the all-in-one manuscript to the main and SI parts.

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage[labelfont=bf, textfont=footnotesize]{caption} % \caption*{} will create legend (explanatory text)
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{lineno}
\linenumbers
\usepackage{float}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{authblk}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{plainnat}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Aquifer depletion exacerbates agricultural drought losses in the US High Plains},
  pdfauthor={Taro Mieno; Timothy Foster; Shunkei Kakimoto; Nicholas Brozovic},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

Two major pathways exist by which aquifer depletion will impact farmers' ability to access and use groundwater as a buffer against drought and rainfall variability \citep{foster2015analysis}. First, depletion increases the energy requirements and costs of pumping water, potentially making it less economically viable for farmers to satisfy crop water needs \citep{mieno2017price, bhattarai2021impact}. Second, depletion also reduces the thickness of an aquifer, lowering the aquifer's transmissivity (i.e., the potential rate of water supply to wells). This latter effect reduces the rate at which water can be pumped and applied to a field as irrigation \citep{konikow2005groundwater, foster2014modeling, hrozencik2017heterogeneous}. If the reduction in well yields is substantial enough - for example as would occur when a large proportion of aquifer thickness is depleted \citep{hecox2002calculation, korus2020depletion} - then a farmer may be unable to extract sufficient water to meet crop water demands, in particular during periods of limited rainfall and in the peak of the growing season when crops are most sensitive to water deficits \citep{foster2015well,rouhi2020downside}. The combined effect of declining well yields and increasing pumping costs could force farmers to reduce irrigation depths, resulting in greater risk of drought-related crop stress. Alternatively, or in addition, a farmer may be forced to limit irrigated area to ensure crop water demands can be met, leading to reductions in total crop production \citep{foster2014modeling, rad2020effects}. Understanding how these two responses vary as a function of aquifer conditions is crucial to guide groundwater management planning decisions, in particular given projected increases in frequency and intensity of droughts under climate change \citep{ukkola2020robust, chiang2021evidence, cook2020twenty}. Another potential pathway is through groundwater use regulation in response to low or declining aquifer thickness. However, it is unlikely that this effects is captured in our analysis. Please see Appendix \ref{pathway-reg}.

@@

Figure \ref{fig:yield-response} shows the estimated crop yield response to water deficit for rainfed production and groundwater-irrigated production at different levels of aquifer thickness categories for corn (panel A) and soybean (panel B). Confidence intervals were not presented to avoid over-crowding the figures. Please see Appendix \ref{reg-conf} for estimated yield response curves with 95 percent confidence interval. The histogram of annual water deficit is presented at the top for each crop. In wetter years with low crop irrigation requirements (i.e., negative values of water deficit), rainfed and irrigated yields are very similar for both corn and soybean. As drought severity increases (i.e., as water deficit increases), rainfed yields of both crops decline significantly whereas irrigated yields are increasing or stable as deficit increases (see Figures \ref{fig:irrigated-yield-ind-corn} and \ref{fig:irrigated-yield-ind-soy} for the 95\% confidence intervals). This result reflects the positive benefits irrigation provides both as a buffer against precipitation variability, and as a means of raising overall yields through increased evapotranspiration and associated biomass accumulation, reductions of heat-related stressors, and other factors \citep{zhu2022untangling, li2020quantifying}.

@@

Figure \ref{fig:ir-share} shows the estimated relationship between aquifer thickness and the share of production area that is irrigated for corn and soybean after accounting for spatio-temporal variations in weather conditions and drought risk exposure across the aquifer. Figure \ref{fig:ir-share} demonstrates that aquifer thickness has meaningfully large impacts on the share of irrigated production for both corn and soybean. As aquifer thickness declines, the share of irrigated acres for corn decreases from around \Sexpr{share_high} (\Sexpr{share_high * 100}\%) for an average county-level aquifer thickness of 100 m to \Sexpr{share_low} (\Sexpr{share_low * 100}\%) for an average county-level aquifer thickness of 10 m (See Appendix \ref{test-dif-share} for formal statistical testings). Reductions in the share of production area under irrigation is also evident for soybean.

@@

where county and year are indicated by subscripts \(i\) and \(t\), respectively, whether rainfed or irrigated is indicated by $ir$,  \(Y\) is crop yield, and \(D\) is seasonal water deficit. \(C_j\) is a dummy variable that takes 1 if observation \(i\) is under category \(C_j\), 0 otherwise. The observations are grouped into four categories: rainfed production, and county-years with aquifer thickness falling in the \([0\%-33.3\%)\), \([33.3\%-66.7\%)\), and \([66.7\% - 100\%]\) quantile ranges of aquifer thickness. The three irrigated production groups have about the same number of observations, which ensures statistical stability in estimating yield response to drought by category. Spline functions are used to capture the potential non-linearity of the impact of water deficit on yield in a flexible manner, with \(\sigma_j\) representing the category-specific intercept, \(\phi_{j,k}\) \(k\)th spline basis function, and \(\beta_k\) is its coefficient (see Appendix \ref{spline-basis} for further explanations on how spline-based regression works). $K$ is the total number of spline basis functions. $K = 3$ was chosen for this regression to prevent our estimated response from becoming unrealistically oscillatory while maintaining a good fit based on Bayesian Information Criterion (BIC). To control for time-invariant heterogeneity across counties and yearly shocks, county-irrigation-status fixed effects $\alpha_{i, ir}$ and year fixed effects $\lambda_t$ are also included. County-irrigation-status fixed effects are used instead of county fixed effects because some counties have both rainfed and irrigated observations, and they need to be distinguished. Standard error estimation is clustered by county.

@@

where \(s_{i,t}\) is the share of irrigated production. The impact of average water deficit is captured by \(\sum_{k}^K \eta_{k}\phi_{k}(D_{i})\), where \(\phi_{k}(\cdot)\) is the \(k\)th spline basis function and \(\eta_{k}\) is its coefficient. The impact of aquifer thickness is captured in a similar way with \(\tau_{l}(\cdot)\) and \(\beta_{l}\) being the \(l\)th spline basis function and its coefficient, respectively. $K$ and $L$ are the numbers of basis functions for $D$ and $A$, respectively. $K = 4$ and $L = 4$ were chosen to balance preventing the estimated responses from being unrealistically oscillatory and ensuring a good fit as indicated by BIC. For this regression, counties with aquifer thickness levels greater than 180 m were excluded due to the sparsity of observations in that range because our flexible functional form approach cannot reliably estimate its impact. County-level soil characteristics including sand percentage, silt percentage, and water holding capacity, are represented by \(X_i\). State-year fixed effects and error term are represented by \(\theta_{s,t}\) and \(v_{i,t}\), respectively. Unlike for our regression models of per-area yields, we do not include county fixed effects to avoid eliminating the majority of variation in the aquifer thickness variable. The vast majority of variation in aquifer thickness is cross-sectional (across counties), rather than over time. Including county fixed effects would effectively eliminate the cross-sectional variation in aquifer thickness. A formal discussion of the trade-offs associated with the choice of fixed effects is provided in Appendix \ref{county-fe}.

@@

% {\centering \includegraphics[width=6in,height=500px,]{../Results/Figures/g_total_impact}}
\caption*{Note: The solid lines represent average yield at different levels of water deficit and aquifer thickness, considering both per-area yield responses and farmers' choices about irrigated area shares in response to different water deficit and aquifer conditions. See Appendix \ref{reg-conf} for estimated average yield response curves with 95\% confidence intervals displayed, which are omitted here for visual clarity. The histograms on the top of each panel show the distribution of county-level annual water deficit values included in regressions for each crop.}
\label{fig:tot-impact}
\end{figure}

@@

\hypertarget{pathway-reg}{%
\section{Groundwater management policies as a potential causal pathway}\label{pathway-reg}}

@@

\setcounter{figure}{0}
\renewcommand{\thefigure}{B.\arabic{figure}}

@@

%------------------------
%- Figure B.1
%------------------------

@@

%------------------------
%- Figure B.2
%------------------------

@@

%------------------------
%- Figure B.3
%------------------------

@@

\setcounter{figure}{0}
\renewcommand{\thefigure}{C.\arabic{figure}}

@@

%------------------------
%- Figure C.1
%------------------------

@@

@@

\setcounter{figure}{0}
\renewcommand{\thefigure}{D.\arabic{figure}}

@@

%------------------------
%- Figure D.1
%------------------------

@@

%------------------------
%- Figure D.2
%------------------------

@@

\setcounter{figure}{0}
\renewcommand{\thefigure}{E.\arabic{figure}}

%------------------------
%- Figure E.1
%------------------------

\label{fig:ex-fx}
\end{figure}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage[labelfont=bf, textfont=footnotesize]{caption} % \caption*{} will create legend (explanatory text)
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{lineno}
% \linenumbers
\usepackage{float}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{authblk}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{plainnat}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\thispagestyle{empty}
\hypersetup{
  pdftitle={Aquifer depletion exacerbates agricultural drought losses in the US High Plains},
  pdfauthor={Taro Mieno; Timothy Foster; Shunkei Kakimoto; Nicholas Brozovic},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

Two major pathways exist by which aquifer depletion will impact farmers' ability to access and use groundwater as a buffer against drought and rainfall variability \citep{foster2015analysis}. First, depletion increases the energy requirements and costs of pumping water, potentially making it less economically viable for farmers to satisfy crop water needs \citep{mieno2017price, bhattarai2021impact}. Second, depletion also reduces the thickness of an aquifer, lowering the aquifer's transmissivity (i.e., the potential rate of water supply to wells). This latter effect reduces the rate at which water can be pumped and applied to a field as irrigation \citep{konikow2005groundwater, foster2014modeling, hrozencik2017heterogeneous}. If the reduction in well yields is substantial enough - for example as would occur when a large proportion of aquifer thickness is depleted \citep{hecox2002calculation, korus2020depletion} - then a farmer may be unable to extract sufficient water to meet crop water demands, in particular during periods of limited rainfall and in the peak of the growing season when crops are most sensitive to water deficits \citep{foster2015well,rouhi2020downside}. The combined effect of declining well yields and increasing pumping costs could force farmers to reduce irrigation depths, resulting in greater risk of drought-related crop stress. Alternatively, or in addition, a farmer may be forced to limit irrigated area to ensure crop water demands can be met, leading to reductions in total crop production \citep{foster2014modeling, rad2020effects}. Understanding how these two responses vary as a function of aquifer conditions is crucial to guide groundwater management planning decisions, in particular given projected increases in frequency and intensity of droughts under climate change \citep{ukkola2020robust, chiang2021evidence, cook2020twenty}. Another potential pathway is through groundwater use regulation in response to low or declining aquifer thickness. However, it is unlikely that this effects is captured in our analysis. For more details on this issue, please refer to section \ref{pathway-reg} of the Supplementary Information.

@@

Figure \ref{fig:yield-response} shows the estimated crop yield response to water deficit for rainfed production and groundwater-irrigated production at different levels of aquifer thickness categories for corn (panel A) and soybean (panel B). Confidence intervals were not presented to avoid over-crowding the figures. Please see section \ref{reg-conf} of the Supplementary Information for estimated yield response curves with 95 percent confidence interval. The histogram of annual water deficit is presented at the top for each crop. In wetter years with low crop irrigation requirements (i.e., negative values of water deficit), rainfed and irrigated yields are very similar for both corn and soybean. As drought severity increases (i.e., as water deficit increases), rainfed yields of both crops decline significantly whereas irrigated yields are increasing or stable as deficit increases (see Figures \ref{fig:irrigated-yield-ind-corn} and \ref{fig:irrigated-yield-ind-soy} for the 95\% confidence intervals). This result reflects the positive benefits irrigation provides both as a buffer against precipitation variability, and as a means of raising overall yields through increased evapotranspiration and associated biomass accumulation, reductions of heat-related stressors, and other factors \citep{zhu2022untangling, li2020quantifying}.

@@

Figure \ref{fig:ir-share} shows the estimated relationship between aquifer thickness and the share of production area that is irrigated for corn and soybean after accounting for spatio-temporal variations in weather conditions and drought risk exposure across the aquifer. Figure \ref{fig:ir-share} demonstrates that aquifer thickness has meaningfully large impacts on the share of irrigated production for both corn and soybean. As aquifer thickness declines, the share of irrigated acres for corn decreases from around \Sexpr{share_high} (\Sexpr{share_high * 100}\%) for an average county-level aquifer thickness of 100 m to \Sexpr{share_low} (\Sexpr{share_low * 100}\%) for an average county-level aquifer thickness of 10 m (Please see section \ref{test-dif-share} of the Supplementary Information for formal statistical testings). Reductions in the share of production area under irrigation is also evident for soybean.

@@

where county and year are indicated by subscripts \(i\) and \(t\), respectively, whether rainfed or irrigated is indicated by $ir$,  \(Y\) is crop yield, and \(D\) is seasonal water deficit. \(C_j\) is a dummy variable that takes 1 if observation \(i\) is under category \(C_j\), 0 otherwise. The observations are grouped into four categories: rainfed production, and county-years with aquifer thickness falling in the \([0\%-33.3\%)\), \([33.3\%-66.7\%)\), and \([66.7\% - 100\%]\) quantile ranges of aquifer thickness. The three irrigated production groups have about the same number of observations, which ensures statistical stability in estimating yield response to drought by category. Spline functions are used to capture the potential non-linearity of the impact of water deficit on yield in a flexible manner, with \(\sigma_j\) representing the category-specific intercept, \(\phi_{j,k}\) \(k\)th spline basis function, and \(\beta_k\) is its coefficient (For further explanations on how spline-based regression functions, please see section \ref{spline-basis} of the Supplementary Information.). $K$ is the total number of spline basis functions. $K = 3$ was chosen for this regression to prevent our estimated response from becoming unrealistically oscillatory while maintaining a good fit based on Bayesian Information Criterion (BIC). To control for time-invariant heterogeneity across counties and yearly shocks, county-irrigation-status fixed effects $\alpha_{i, ir}$ and year fixed effects $\lambda_t$ are also included. County-irrigation-status fixed effects are used instead of county fixed effects because some counties have both rainfed and irrigated observations, and they need to be distinguished. Standard error estimation is clustered by county.

@@

where \(s_{i,t}\) is the share of irrigated production. The impact of average water deficit is captured by \(\sum_{k}^K \eta_{k}\phi_{k}(D_{i})\), where \(\phi_{k}(\cdot)\) is the \(k\)th spline basis function and \(\eta_{k}\) is its coefficient. The impact of aquifer thickness is captured in a similar way with \(\tau_{l}(\cdot)\) and \(\beta_{l}\) being the \(l\)th spline basis function and its coefficient, respectively. $K$ and $L$ are the numbers of basis functions for $D$ and $A$, respectively. $K = 4$ and $L = 4$ were chosen to balance preventing the estimated responses from being unrealistically oscillatory and ensuring a good fit as indicated by BIC. For this regression, counties with aquifer thickness levels greater than 180 m were excluded due to the sparsity of observations in that range because our flexible functional form approach cannot reliably estimate its impact. County-level soil characteristics including sand percentage, silt percentage, and water holding capacity, are represented by \(X_i\). State-year fixed effects and error term are represented by \(\theta_{s,t}\) and \(v_{i,t}\), respectively. Unlike for our regression models of per-area yields, we do not include county fixed effects to avoid eliminating the majority of variation in the aquifer thickness variable. The vast majority of variation in aquifer thickness is cross-sectional (across counties), rather than over time. Including county fixed effects would effectively eliminate the cross-sectional variation in aquifer thickness. A formal discussion of the trade-offs associated with the choice of fixed effects is provided in section \ref{county-fe} of the Supplementary Information.

@@

% {\centering \includegraphics[width=6in,height=500px,]{../Results/Figures/g_total_impact}}
\caption*{Note: The solid lines represent average yield at different levels of water deficit and aquifer thickness, considering both per-area yield responses and farmers' choices about irrigated area shares in response to different water deficit and aquifer conditions. Please see section \ref{reg-conf} of the Supplementary Information for estimated average yield response curves with 95\% confidence intervals displayed, which are omitted here for visual clarity. The histograms on the top of each panel show the distribution of county-level annual water deficit values included in regressions for each crop.}
\label{fig:tot-impact}
\end{figure}

@@

% \renewcommand{\thesection}{S.I. \arabic{section}}
\setcounter{figure}{0}
\renewcommand{\thefigure}{S.\arabic{figure}}

\hypertarget{pathway-reg}{%
\section{}\label{pathway-reg}}

@@

@@

%------------------------
%- Figure S.1
%------------------------

@@

%------------------------
%- Figure S.2
%------------------------

@@

%------------------------
%- Figure S.3
%------------------------

@@

@@

%------------------------
%- Figure S.4
%------------------------

@@

%------------------------
%- Figure S.5
%------------------------

%------------------------
%- Figure S.6
%------------------------

\label{fig:ex-fx}
\end{figure}
\clearpage

@@

@@

%------------------------
%- Figure S.7
%------------------------

@@

%------------------------
%- Figure S.8
%------------------------

@@
